---
title: "Airlines"
author: "Ben Baumer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Airlines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE}
library(dplyr)
library(airlines)
```

## Create the airlines database

First, we create a connection to the database server. We can work with database connections used by either `dplyr` or `DBI`. 

```{r}
library(RPostgreSQL)
#' # must have pre-existing database "airlines"
db <- src_postgres(host = "localhost", user = "postgres", password = "postgres", dbname = "airlines")
```

```{r, eval=FALSE}
# library(RMySQL)
#' # must have pre-existing database "airlines"
# db <- src_mysql(host = "localhost", user = "mdsr", password = "mdsr", dbname = "airlines")
```


Next, download `flights` for a single year, and push it to the database server. 

```{r, eval=FALSE}
pushFlightsYear(db, year = 2013, temp.dir = "~/dumps/airlines")
```

Given more time, we might loop through many years. 

```{r, eval=FALSE}
years <- 2014:2015
lapply(years, pushFlightsYear, db = db, temp.dir = "~/dumps/airlines")
```

There are several ancillary tables that we can also push to the database server. 

```{r, eval=FALSE}
# copy_to(db, carriers, indexes = list("carrier"), temporary = FALSE)
dbWriteTable(db$con, "carriers", as.data.frame(carriers), overwrite=TRUE, row.names = FALSE)
dbWriteTable(db$con, "airports", as.data.frame(airports), overwrite=TRUE, row.names = FALSE)
dbWriteTable(db$con, "planes", as.data.frame(planes), overwrite=TRUE, row.names = FALSE)
dbWriteTable(db$con, "weather", as.data.frame(weather), overwrite=TRUE, row.names = FALSE)
```

To complete the setup, we should build primary keys and indexes on these tables. 

```{r, eval=FALSE}
buildIndices(db)
```

## Accessing the airlines database

Let's see what has been created:

```{r}
dbListTables(db$con)
dbListFields(db$con, "airports")
```

Now we can connect to these tables using `dplyr`:

```{r}
airports <- tbl(db, "airports")
planes <- tbl(db, "planes")
carriers <- tbl(db, "carriers")
flights <- tbl(db, "flights")
```

We can retrieve some basic information about what results are present in the database. 

```{r, eval=FALSE}
flights %>%
   group_by(year, origin) %>%
   summarise(N = n(), numDests = count(distinct(dest)), 
             numCarriers = count(distinct(carrier)), 
             numPlanes = count(distinct(tailnum))) %>%
   arrange(desc(N))
```

## Recover nycflights13

To restrict this to only flights to and from the three New York City airports in 2013, we simply `filter` and `trim`:

```{r, eval=FALSE}
nycFlights13 <- flights %>%
  filter(year == 2013) %>%
  filter(origin %in% c("JFK", "LGA", "EWR"))

tbl_list <- trim(db, flights = nycFlights13)
airports <- collect(tbl_list$airports)
# save(airports, file = "data/airports.rda")
```