---
title: "Airlines"
author: "Ben Baumer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Airlines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Install packages

```{r, eval=FALSE, message=FALSE}
install.packages("devtools")
devtools::install_github("beanumber/etl")
devtools::install_github("beanumber/airlines")
```

```{r, message=FALSE}
library(airlines)
```

## Create the airlines database

First, we create a connection to the database server. We can work with database connections used by either `dplyr` or `DBI`. 

```{r}
# library(RPostgreSQL)
# must have pre-existing database "airlines"
# db <- src_postgres(host = "localhost", user = "postgres", dbname = "airlines")
```

```{r, eval=FALSE}
library(RMySQL)
# must have pre-existing database "airlines"
db <- src_mysql(host = "localhost", user = "r-user", password = "mypass", dbname = "airlines")
```

```{r, eval=FALSE}
airlines <- etl("airlines", db, dir = "~/dumps/airlines")
```

If the database has not been populated yet, we use the `etl_create` function to initialize it and insert some data. 

```{r, eval=FALSE}
library(magrittr)
airlines %<>%
  etl_create(year = 1987, months = 10)
```

Once the database is set up, you can update the flights table with more data.  

```{r, eval=FALSE}
airlines %<>%
  etl_update(year = 1987, months = 11:12) %>%
  etl_update(year = 1988)
```

Given more time, we might loop through many years. **NOTE: this will take a looong time.**

```{r, eval=FALSE}
lapply(1989:2015, etl_update, obj = airlines)
```

If the process get disturbed, you can fine-tune using the individual [ETL](https://github.com/beanumber/etl/blob/master/README.md) functions. Note `etl_update` is just a wrapper for these functions.

```{r}
getS3method("etl_update", "default")
```

Furthermore, `etl_create` is just a wrapper for `etl_update`.

```{r}
getS3method("etl_create", "default")
```

## Accessing the airlines database

Let's see what has been created:

```{r, eval=FALSE}
dbListTables(db$con)
dbListFields(db$con, "airports")
```

Now we can connect to these tables using `dplyr`:

```{r, eval=FALSE}
airports <- tbl(db, "airports")
planes <- tbl(db, "planes")
carriers <- tbl(db, "carriers")
flights <- tbl(db, "flights")
```

We can retrieve some basic information about what results are present in the database. 

```{r, eval=FALSE}
flights %>%
   group_by(year, origin) %>%
   summarise(N = n(), numDests = count(distinct(dest)), 
             numCarriers = count(distinct(carrier)), 
             numPlanes = count(distinct(tailnum))) %>%
   arrange(desc(N))
```

## Recover nycflights13

To restrict this to only flights from the three New York City airports in 2013, we simply `filter` and `trim`:

```{r, eval=FALSE}
nycFlights13 <- flights %>%
  filter(year == 2013) %>%
  filter(origin %in% c("JFK", "LGA", "EWR"))

tbl_list <- trim(db, flights = nycFlights13)
airports <- collect(tbl_list$airports)
# save(airports, file = "data/airports.rda")
```

Verify that the data seems accurate.

```{r, eval=FALSE}
flights <- db %>%
  tbl(from = "flights")
flights %>%
  summarise(numFlights = n())
```

This should return about 163 million flights from October 1987 to June 2015. 

```{r, eval=FALSE}
flights %>%
#  filter(year > 2003) %>%
  group_by(year) %>%
  summarise(numMonths = count(distinct(month)), numFlights = n()) %>%
  print(n = 40)
```

Check the size of your partiions.

```{bash, eval=FALSE}
sudo ls -lhS /var/lib/mysql/airlines/ | grep .MYD
```